\LARGE \bf Thymio, an example of open source hardware robot \\ for formal education

### 2016-03-31 paper_ram.tex
improved version

There is a very large number of publications presenting educational robots with different prices and features, from very low-cost systems targeting, for instance, education in Africa~\cite{Rubenstein2015,Gyebi2015} to extremely sophisticated humanoids~\cite{Hood2015,Mazzoni2016}.
Among those, only very few are commercially available, limiting their validation by education scientists.
This results in 90\% of publications about validation of educational results made on LEGO Mindstorm~\cite{benitti2012exploring}.
The LEGO Mindstorm EV3\footnote{\url{http://mindstorms.lego.com}} is expensive ($\sim$400\,\$) but offers a wide range of possibilities, especially at the mechanical level using LEGO bricks and at the software level with its graphical programming environment. 
At a lower price, LEGO proposes the WeDo system ($\sim$80\,\$).
The WeDo is similar to EV3 but only reads one sensor and controls one actuator and must always be connected to a desktop computer.
The BeeBot robot\footnote{\url{http://www.beebot.org.uk}} is a cheap ($\sim$90\,\$) autonomous system.
However, it is not a real robot but rather an automaton: There is no perception of the environment, just a good encoder for precise movements and a simple programming interface based on buttons giving directions for the movements.

Among these widely used robots, none is part of an open source hardware project.
The only effort in this direction comes from LEGO, that made public all the schematics of the EV3 in an hardware developer kit, to allow the design of extensions. 
Otherwise, only few educational robots are both used in schools and are issued from an open hardware project:
Scribbler2, produced and sold by Parallax\footnote{\url{http://www.parallax.com}} ($\sim$180\,\$), is a 188~mm large robot, designed to run on the ground and equipped with few light sensors, one distance sensor, two ground sensors and few LEDs display.
It runs on standard AA batteries and has a hacker port for interfacing electronic extensions.
It is programmable with a graphic or text code interface.  
The main weakness of Scribbler2 is its performance per price ratio and the limited compatibility with other systems. 
The e-puck~\cite{mondada2009puck} robot is a robot targeting university-level education.
Well equipped with sensors and actuators, modular and compact, it can be programmed with professional environments.
Several simulators allow to run experiments of high complexity. 
It's main weakness is its high price ($\sim$870\,\$).
Finch\footnote{\url{http://www.finchrobot.com/}} ($\sim$99\,\$) is a very simple robot that has been designed around a wired connection to the computer. 
This connection reduces electronics requirements, such as batteries or wireless communication, and allows to implement the control entirely on the computer.
This results in a very broad set of possible programming languages available, the real force of this robot. 
On the other hand the cable does not allow real autonomy and mobility of the robot.
Finally the mBot\footnote{\url{http://www.makeblock.cc/mbot/}} is a mobile platform based on the Arduino board. 
The very cheap and simple electronics allow to drastically reduce the price of the system ($\sim$75\,\$) but also reduces the perception possibilities to a couple of sensors.

@@

@@

The choice of developing a robot within an open source hardware project has an impact on the robot design and the way this device is used in the community of users. 
In this section we analyze more in detail the implications of this choice in the context of educational robotics.
We compare these implications with the result of a survey we did among 25 people active in various open source hardware projects based on a worldwide call for contributions. 
Among these 25 people, 6 are project leaders, 10 are part of the core design team, 6 are contributors and 3 are enthusiastic users.
60\% of these respondents are between 25 and 35 years old and 80\% either work in an academic environment or are students.

@@

When asked about their motivation to participate to open source hardware projects, the people participating to the survey give as main motivation a link to their professional activity and to the specific project, followed by a more general motivation by the nature of the project and finally by the engineering practice. 
The more abstract goal of improving our society is mentioned only by few respondents.

\begin{figure}
\centering
\includegraphics[width=.7\columnwidth]{figures/motivation}
\caption{Motivation of the people participating to the survey.}
\label{fig:motivation}
\end{figure}

@@

It is more interesting to look at what people expect as benefits in participating in such a project.
Most of the answers mention impact, networking and learning. 
In these answers we can observe both a technological component in getting better results and impact, and a human-relations component that is not in the basic definition of open hardware but results from the community created around the project. 
People expect to work with other people to achieve more together, to create a network of competences or users, to learn through new experience and contacts.

@@

\begin{figure}
\centering
\includegraphics[width=.7\columnwidth]{figures/getout}
\caption{Motivation of the people participating to the survey.}
\label{fig:getout}
\end{figure}

@@

There is another licensing issue which is not well known and that we discovered very late in our project: the constrains of the license of the mechanical and electronic CAD tools. 
Indeed, asked about this issue, the participants to our survey seem generally not aware of the fact that CAD licenses can be very restrictive about the way the source files can be published.
Around the half or respondents to our survey, also among the project leaders, state that they were not aware that this could be an issue, and only one third checked the license of their CAD software. 
Among the respondents, some stated that they did not checked the license because they were sure they can freely publish their design, but half of them are using software not allowing the publication of source files when using educational licenses, and all these respondents are academic or students.
This issue is very serious, as most contributors of open source hardware projects are academic and use academic or educational licenses. 
Most licenses do not allow to use the design created with this version of the software for commercial activities. 
As producing or selling the product is part of the definition of open hardware, these licenses simply forbid publication under the standard open source hardware conditions.

\begin{figure}
\centering
\includegraphics[width=.9\columnwidth]{figures/aware-license}
\caption{Awarness about CAD tool licenses limitations by people participating to the survey.}
\label{fig:aware}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/table}
\caption{Publication possibilities as function of the CAD editors, situation in February 2016.}
\label{fig:editors}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/production}
\caption{Production methods in the projects mentioned by the people participating to the survey.}
\label{fig:production}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/problems}
\caption{Problems observed by participants to open source hardware projects.}
\label{fig:problems}
\end{figure}

The very specific structure of an open project has also drawbacks. 
Asked about problems met in their project, the participants to our survey mention the management of the community, the evolution of the project, the efforts in documentation, production and distribution and the limited resources.
There are here two kind of problems: one of management of people, the other in the development of the project after the first design.
The management of the project seems a recurrent problem and is related to the nature of this type of structure, mainly very flat and built on academic people with engineering background. 
Most of the projects, following our survey, are managed by the founder of the initiative and not by the most expert in the field or by a manager, which is perhaps not the best objective choice.
It is indeed very difficult to evolve from a phase where the engineering is the core of the project, to a new phase where communication and management play an increasing role.
The same holds for several other factors, like the efforts in documentation, production and distribution of the product. 
Also here there is a change of activity from the very creative and appealing design phase, to a phase of documentation and, in some cases, production and distribution which is not at all the same type of activity and requires different skills.
This shift requires extra efforts and motivation, which are often missing, especially for activities like documentation that are not appreciated by engineers.
In general many respondents to our survey identified the evolution of the project as a general problematic issue, which summarizes the several aspects mentioned above.

@@

XXXXXXX historical, should we keep this?

We reached our price target thanks to our careful selection of components and our production in China.
Thymio costs 35\,\$ per unit to produce in batch of 1\,k units (see \ref{tbl:thymio-price}), with fixed fees of 15\,k\,\$ (13\,k\,\$ for the mold and 2\,k\,\$ for the \textsc{ce} certification).
This allows a selling price of 100\,CHF ($\approx$ 106\,\$).
This small ratio of selling-over-production price is made possible by our non-profit business model.

XXXXXXX end of historical

@@

>>>>

There is a very large number of publications presenting educational robots with different prices and features, from very low-cost systems targeting, for instance, education in Africa~\cite{Rubenstein2015,Gyebi2015} to extremely sophisticated humanoids~\cite{Hood2015,Mazzoni2016}.
Among those, only very few are commercially available, limiting their validation by education scientists.
This results in 90\% of publications about validation of educational results made on LEGO Mindstorm~\cite{benitti2012exploring}.
The LEGO Mindstorm EV3\footnote{\url{http://mindstorms.lego.com}} is expensive ($\approx$ 400\,\$) but offers a wide range of possibilities, especially at the mechanical level using LEGO bricks and at the software level with its graphical programming environment. 
At a lower price, LEGO proposes the WeDo system ($\approx$ 80\,\$).
The WeDo is similar to EV3 but only reads one sensor and controls one actuator and must always be connected to a desktop computer.
The BeeBot robot\footnote{\url{http://www.beebot.org.uk}} is a cheap ($\approx$ 90\,\$) autonomous system.
However, it is not a real robot but rather an automaton: There is no perception of the environment, just a good encoder for precise movements and a simple programming interface based on buttons giving directions for the movements.

Among these widely used robots, none is part of an open source hardware project.
The only effort in this direction comes from LEGO, that made public all the schematics of the EV3 in an hardware developer kit, to allow the design of extensions. 
Otherwise, only few educational robots are both used in schools and are issued from an open hardware project:
Scribbler2, produced and sold by Parallax\footnote{\url{http://www.parallax.com}} ($\approx$ 180\,\$), is a 188~mm large robot, designed to run on the ground and equipped with few light sensors, one distance sensor, two ground sensors and few LEDs display.
It runs on standard AA batteries and has a hacker port for interfacing electronic extensions.
It is programmable with a graphic or text code interface.  
The main weakness of Scribbler2 is its performance per price ratio and the limited compatibility with other systems. 
The e-puck~\cite{mondada2009puck} robot is a robot targeting university-level education.
Well equipped with sensors and actuators, modular and compact, it can be programmed with professional environments.
Several simulators allow to run experiments of high complexity. 
It's main weakness is its high price ($\approx$ 870\,\$).
Finch\footnote{\url{http://www.finchrobot.com/}} ($\approx$ 99\,\$) is a very simple robot that has been designed around a wired connection to the computer. 
This connection reduces electronics requirements, such as batteries or wireless communication, and allows to implement the control entirely on the computer.
This results in a very broad set of possible programming languages available, the real force of this robot. 
On the other hand the cable does not allow real autonomy and mobility of the robot.
Finally the mBot\footnote{\url{http://www.makeblock.cc/mbot/}} is a mobile platform based on the Arduino board. 
The very cheap and simple electronics allow to drastically reduce the price of the system ($\approx$ 75\,\$) but also reduces the perception possibilities to a couple of sensors.

@@

Thymio finally costs 35\,\$ per unit when produced in a batch of 1\,k units (see \ref{tbl:thymio-price}), with fixed fees of 15\,k\,\$ (13\,k\,\$ for the mold and 2\,k\,\$ for the \textsc{ce} certification).
This allows a selling price of 129\,CHF ($\approx$ 130\,\$).

@@

The choice of developing a robot within an open source hardware project has an impact on the robot design and the way this device is used in the community of users. 
In this section we analyze more in detail the implications of this choice in the context of educational robotics.
We compare these implications with the result of a survey that got 35 answers from people active in various open source hardware projects based on a worldwide call for contributions. 
Among these 35 answers, 11 come from project leaders, 13 from core design team members, 8 from contributors and 3 from enthusiastic users.
54\% of these respondents are between 25 and 35 years old, 26\% between 35 and 50 years old, 9\% are less than 25 years old and 11\% are older than 50.  
67\% either work in an academic environment or are students. 
34 of the 35 answers come from male repondents.
This last point characterizes strongly the hardware developers community.

@@

When asked about their motivation to participate to open source hardware projects, the people participating to the survey give as main motivation a link to their professional activity and to the specific project, followed by a more general motivation by the nature of the project and finally by the engineering practice (see Figure~\ref{fig:motivation}). 
The more abstract goal of improving our society is mentioned only by few respondents.

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/motivation}
\caption{Motivation of the people participating to the survey.}
\label{fig:motivation}
\end{figure}

@@

It is more interesting to look at what people expect as benefits in participating in such a project (see Figure~\ref{fig:getout}).
Most of the answers mention impact, networking and learning. 
In these answers we can observe both a technological component in getting better results and impact, and a human-relations component that is not in the basic definition of open hardware but results from the community created around the project. 
People expect to work with other people to achieve more together, to create a network of competences or users, to learn through new experience and contacts.

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/advantages}
\caption{Advantages that people participating to the survey perceive by contributing to an open source hardware project.}
\label{fig:getout}
\end{figure}

@@

There is another licensing issue which is not well known and that we discovered very late in our project: the constrains of the license of the mechanical and electronic CAD tools. 
Indeed, asked about this issue, the participants to our survey seem generally not aware of the fact that CAD licenses can be very restrictive about the way the source files can be published (see Figure~\ref{fig:aware}).
More than the half or respondents to our survey, and more than 60\% among the project leaders, state that they were not aware that this could be an issue, and only one third checked the license of their CAD software. 
Among the respondents, some stated that they did not checked the license because they were sure they can freely publish their design, but one third of them are using software not allowing the publication of source files when using educational licenses, and all these respondents are academic or students.
This issue is very serious, as most contributors of open source hardware projects are academic and use academic or educational licenses. 
Most licenses do not allow to use the design created with this version of the software for commercial activities. 
As producing or selling the product is part of the definition of open hardware, these licenses simply forbid publication under the standard open source hardware conditions.

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/checklicense}
\caption{Awarness about CAD tool licenses limitations by people participating to the survey.}
\label{fig:aware}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/table}
\caption{Publication possibilities as function of the CAD editors, situation end of March 2016.}
\label{fig:editors}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/manufacturing}
\caption{Production methods in the projects mentioned by the people participating to the survey.}
\label{fig:production}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/limitations}
\caption{Limitations of open source hardware projects observed by participants to open source hardware projects.}
\label{fig:problems}
\end{figure}

The very specific structure of an open project has also drawbacks. 
Asked about problems met in their project (see Figure~\ref{fig:problems}), the participants to our survey mention the management of the community, the evolution of the project, the efforts in documentation, production and distribution and the limited resources.
There are here two kind of problems: one of management of people, the other in the development of the project after the first design.
The management of the project seems a recurrent problem and is related to the nature of this type of structure, mainly very flat and built on academic people with engineering background. 
Most of the projects, following our survey, are managed by the founder of the initiative and not by the most expert in the field or by a manager, which is perhaps not the best objective choice.
It is indeed very difficult to evolve from a phase where the engineering is the core of the project, to a new phase where communication and management play an increasing role.
The same holds for several other factors, like the efforts in documentation, production and distribution of the product. 
Also here there is a change of activity from the very creative and appealing design phase, to a phase of documentation and, in some cases, production and distribution which is not at all the same type of activity and requires different skills.
This shift requires extra efforts and motivation, which are often missing, especially for activities like documentation that are not appreciated by engineers.
In general many respondents to our survey identified the evolution of the project as a general problematic issue, which summarizes the several aspects mentioned above.

@@

@@

The open source hardware development and dissemination of Thymio addresses several of the issues found in educational robotics.
It allowed to broadly distributed the robot with minimal changes dues to management of intellectual properties, royalties, financial support and so on.
This was achieved with an excellent match between the philosophy of the project and the one of the community of users in education.
In particular, the open source approach allow to provide a durable robot, easy to maintain and repair, with at the same time a community of users providing educational material and mutual support.
